Probabilistic Topic Modelling for Spatial Smoothing in Mass Spectrometry Imaging

### 2017-01-20 docs/msci-proposal/mprop.tex
Renaming the MSci project.

\par The challenges on working with MSI data arise from the data capturing process. In metabolomics, the MS equipment requires the ionisation of metabolites. Effectively, a metabolite is compounded with an appropriate chemical entity to produce an ion. It follows that the mass spectrometers are capable to capture some properties of the ionised metabolites. Since the mass of an ion belongs to these properties, the collection of captured ions can be represented in terms of mass spectra. However, the following issues have to be considered upon reasoning from the MSI data: (1) the ionisation of a metabolite can be performed using different chemical entities, therefore the ions of same metabolite type would have different masses; (2) a metabolite might be fragmented during the ionisation; (3) the metabolites of same type might have different atom structure, which suggests a different mass value. The latter issues, effectively, corresponds to a problem of error correction.

@@

% Unsupervisder learning
% - Topic moels
% - Probabilistic models
% - LDA:
% -- General
% -- Metabolomics
\par Since this proposal is focused on utilising spatial smoothing to reduce the impact of noise in MSI data sets, we will familiarise with the machine learning techniques and problems tackled in the MSI domain. The objective to discover novel patterns can be treated as a problem of \textit{unsupervised} machine learning. That is, we have no initial settings in how the patterns are expected to look like. However, upon appropriate optimisation, the unsupervised machine learning techniques are capable to learn the patterns on their own. To be more specific, we will utilise the techniques of \textit{topic modelling} (in this case, a pattern is referred by a topic). Also, we are focusing on the \textit{probabilistic} (statistical) models, more specifically, \textit{latent Dirichlet allocation} (LDA). First of all, the successor LDA-like models are treated as a state-of-the-art method in tackling semantic analysis type problems. For more detailed review on LDA applications, the reader can consult the survey conducted by Blei \cite{blei_2012}. Also, Hooft et al. \cite{hooft} has recently addressed the applicability of LDA in MSI data sets. This proposal is directly influenced by the latter study: we will utilise the LDA-like models to reduce the impact of noise in the MSI data sets.

% Outline
% - Sections
\par The structure of this proposal is given as follows: in Section 2 we present the issue of noise in MSI data sets; in section 3 we cover the literature in topic modelling which focuses on the variations of LDA-like models; in Section 4 we discuss the approach to tackle the problem (we discuss the reviewed literature, address design and evaluation details); finally, in Section 5, we set and schedule the tasks required to complete the project.
% To-do: rewrite the outline

@@

% Experiments
% - Document modelling: issue with pLSI and mixture unigrams~ 
% - Document classification: 
% -- Dimensionality reduction
% -- Comparison with SVM
\par The conducted experiments display an improvement in performance over the predecessor models and suggest the application domains of LDA-like models. The first experiment addresses document modelling. To be more specific, LDA has displayed balanced hidden topic proportions in the set of tested documents. That is, the tested documents have replicated the topic proportions of the training documents. The second experiment is directed towards the problem of document classification. The authors have addressed the \textit{dimensionality reduction} in order to surpass the performance of the support vector machine (SVM) models. To be more specific, the documents with reduced dimensionality would possess lower amount of features. The results of the experiment indicate that the basic LDA model has successfully reduced the number of features. Also, it has displayed an improvement in accuracy compared to the SVM-like models.

@@

% Inference
% Variational methods are a deterministic alternavive for stochastic simulation
% Two variations: Kalman filtering and wavelet regression
\par The authors have used variational methods to approximate the inference of the posterior. It is claimed that stochastic simulation (sampling-based variance methods) would be unable to scale with larger data sets. Further, the authors discuss two methods used for the variational inference: Kalman Filtering and Wavelet Regression. The implementation details of these methods can be found in the `Dynamic Topic Modelling' article \cite{blei_2006}. Note that the implementation details are provided in a brief manner suggesting the need of supplementary literature.

@@

The initial dynamic topic modelling paper suggests to induce the time-series into the generative process of data sets. However, the way we capture the ions is done instantaneously for the whole corpus. Therefore, we are relaxing the assumption of time-series. Instead of that, we will use the dynamic topic modelling techniques to establish the relationship between the time-slices (even though we are not taking time into account, we will stick to the same terminology to preserve the consistency). For example, if we assume that the data is generated in a 1-dimensional manner, then time-slice $t$ would depends on its adjacent time-slices $t-1$ and $t+1$. However, since the previous illustration of the sample is 2-dimensional, there are more adjacent time-slices. This relationship is shown in Figure \ref{fig:2-d_generative_process} below.
\begin{figure}[H]
  \centering
  \includegraphics[width=0.5\textwidth]{2-d_generative_process}
  \caption{2-dimensional generative process.}
  \label{fig:2-d_generative_process}
\end{figure}
In the boundary case shown on the left, time-slice $t_{1,1}$ is related to three surrounding time-slices, whereas, in the general case indicated on the right side, time-slice $t_{2,2}$ relates to eight surrounding time-slices. Ultimately, the notion of dynamic topic modelling can be  preserved even upon relaxing the assumption of time-series.

@@

% Inference
% - Metropolis--Hastings
\par In Section 3, we have addressed a sampling-based inference method -- Gibbs sampling. However, for the dynamic topic models, the latter method has to be reconsidered due to the change from Dirichlet to Gaussian distributions. Therefore, we will look into another MCMC based algorithm -- Metropolis--Hastings.
% - To-do: Analyse the details of the MH.

@@

% Intro
% - Spatial Smoothing
% - Two inference methods
% - Risk management
\par The developed technique addressing the spatial smoothing of the MSI data sets will evaluated by measuring the improvement in performance. That is, the performance will be assessed by investigating the quality of the inferred patterns. Also, we will utilise two different inference methods: sampling-based and variational. By having two model variations, we will be able to compare the time performance. Hence, we will be able to deduce which variation scales better with the size of the data sets. Finally, the section will be finished by investigating the possible risks; also, we will provide potential strategies mitigating the risks.

% 
% - Generating synthetic data sets
% -- The improvement in the inference
% 
\par The implemented model will be evaluated using synthetic and real data sets. However, at the start, we will use the synthetic data sets. Note that the performance will be measured in the following way: (1) take two equivalent data sets; (2) apply the spatial smoothing on one of the data sets; (3) infer the underlying structure of the data; (4) check whether the applied smoothing has inferred the underlying structures better. By working with synthetic data sets, we had the control over inducing the underlying structure of the data. Hence, we will able to give a basis in measuring the performance of the applied spatial smoothing.

% Risks
% - Algorithm complexity (implementation difficulty)
% - Size of the data sets -> slow computation 
% - Obtaining real data
% - Uncertainty whether the model produces an improvement
\par Even though we have induced a strategy for the project's evaluation, the described process assumes that we can mitigate the potential risks. Note that, at the initial stage of the project, we have identified the following potential risks:
\begin{itemize}
  \item The implementation complexity of the inference algorithms;
  \item The access to the real data.
  \item The model's scalability in large data sets;
  \item The inability to measure the performance of the applied spatial smoothing.
\end{itemize}
The first three risks would slow down the project execution thus raise a concern of finishing the project on time, whereas the fourth one would prevent the confirmation of denial of the initial project's hypothesis -- `the degeneracies of MSI data sets can be detected using LDA-like models'.

% Risk mitigation
% - Algorithms
% -- Familiarising with them before the research
% -- Starting with the simpler one (Gibbs sampling)
\par The risk of complex (in terms of development) algorithms has been already mitigated by familiarising with basic sampling-based and optimisation-based inference models. That is, we have implemented the models utilising Gibbs sampling and variational inference. Also, note that the main model will firstly be developed using Gibbs sampling, as this method is easier to implement compared to the variational one.

@@

% - Real data can be obtained from the Web
\par The risk of the real data access might be partially resolved, as the University has a laboratory for the research in Metabolomics. However, in case this options is not possible, we can obtain the data sets from the Web.

% - Uncertainty in the ground truth
% -- Generate data sets with underlying structure
% -- Assess the inference by using spatial and non-spatial models
\par Finally, the risk of the inability to measure the performance of the applied smoothing can impact the outcome of the project the most. That is, it would be difficult to reason about the improvement in the inference of the real data sets, because we do not possess the full knowledge of the data set underlying structure. We will mitigate this issue by establishing a performance metric from the synthetic data sets.

% Evaluation/Experiments
% - Real data
\par By having the defined boundaries of the expected improvement in performance, we can attempt to tackle the real data sets. That is, the expected performance improvement would serve as an indicator whether the inferred real data deviates in the expected limit. To be more specific, the data sets with applied spatial and the data sets without applied smoothing will display some differences in the inferred topics. Hence, we will check whether those differences meet the performance boundaries suggested by the inference of the synthetic data sets.

@@

% Intro
% - Tasks.
% - Schedule
% - Deliverables
\par In this section, we will outline the project milestones and deliverables. That is, we will outline the most important steps and provide the estimations in a form of Gantt chart. Further, the deliverables will be directly induced from the project's objectives and hypothesis (these are addressed in Section 2). The deliverables will contain both successful and successful outcomes of the set hypothesis.

@@

% Tasks
% - Design
% -- Investigate the literature on the spatial smoothing application in MSI data sets
% -- Design the dynamic topic model
% - Implementation
% -- Implement the generative process of the dynamic topic modelling with applied spatial smoothing
% -- Implement both sampling-based and variational inference to the dynamic topic model
% - Evaluation
% -- Test the performance of the applied spatial smoothing
% - Documentation/Presentation
% -- Dissertation
% -- Reports
% -- Presentation
\par The high-level plan of the project's execution include the following milestones: design, implementation, evaluation, and documentation/presentation. To establish a basis for the design, we will continuously review the literature on spatial smoothing, MSI, and dynamic topic models. Then, the implementation will require the utilisation of spatial smoothing in the generative process of the dynamic topic model; also, for the inference, we will implement both: sampling-based and variational methods. Further, the evaluation will be conducted by assessing the performance of the developed models. Finally, the results will be documented in a form of weekly reports, dissertation, and presentation.

% Dates
% - Non-continuous
% - Estimations
\par The outlined tasks will not be necessarily executed in sequential manner. For some of the tasks, we will use the parallel approach as given in Figure \ref{fig:gantt} below.

@@

% - Successful/Unsuccessful
% - Model in Python 
% -- 2/3 objectives
% - Dissertation
% - Presentation
\par The deliverables of the project will depend on the outcome of the initial hypothesis -- `the degeneracies of MSI data sets can be detected using LDA-like models'. Nevertheless, in any of the project's outcomes, we will provide the following deliverables: the presentation; the dissertation; and the spatial smoothing model, which will be written in Python. However, the main project's contribution to the research in MSI will depend on the results: if the model displays an improvement in performance, then we could utilise the model in pre-processing MSI data sets; in the case of unsuccessful outcome, we would assess the techniques used and provide details on why the model does not work. Both outcomes would bring the motivation for further research in improving MSI data sets: the successful outcome would bring attention to the applicability of LDA-like models; the unsuccessful outcome would suggest the need of a different perspective required to tackle the issue of noise.

>>>>

\par The challenges on working with MSI data arise from the data capturing process. In metabolomics, the MS equipment requires the ionisation of metabolites. Effectively, a metabolite is compounded with an appropriate chemical entity to produce an ion. It follows that the mass spectrometers are capable of capturing some properties of the ionised metabolites. Since the mass of an ion belongs to these properties, the collection of captured ions can be represented in terms of mass spectra. However, the following issues have to be considered upon reasoning from the MSI data: (1) the ionisation of a metabolite can be performed using different chemical entities, therefore the ions of same metabolite type would have different masses; (2) a metabolite might be fragmented during the ionisation; (3) the metabolites of same type might have different atom structure, which suggests a different mass value. The latter issues, effectively, corresponds to a problem of error correction.

@@

% Unsupervisder learning
% - Topic moels
% - Probabilistic models
% - LDA:
% -- General
% -- Metabolomics
\par Since this proposal is focused on utilising spatial smoothing to reduce the impact of noise in MSI data sets, we will familiarise with the machine learning techniques and problems tackled in the MSI domain. The objective to discover novel patterns can be treated as a problem of \textit{unsupervised} machine learning. That is, we have no initial settings in how the patterns are expected to look like. However, upon appropriate optimisation, the unsupervised machine learning techniques are capable of learning the patterns on their own. To be more specific, we will utilise the techniques of \textit{topic modelling} (in this case, a pattern is referred by a topic). Also, we are focusing on the \textit{probabilistic} (statistical) models, more specifically, \textit{latent Dirichlet allocation} (LDA). First of all, the successor LDA-like models are treated as a state-of-the-art method in tackling semantic analysis type problems. For a more detailed review of LDA applications, the reader can consult the survey conducted by Blei \cite{blei_2012}. Also, Hooft et al. \cite{hooft} has recently addressed the applicability of LDA in MSI data sets. This proposal is directly influenced by the latter study: we will utilise the LDA-like models to reduce the impact of noise in the MSI data sets.

% Outline
% - Sections
\par The structure of this proposal is given as follows: in Section 2 we present the issue of noise in MSI data sets; in section 3 we cover the literature in topic modelling which focuses on the variations of LDA-like models; in Section 4 we discuss the approach tackling the noise reduction problem (we discuss the reviewed literature, address design and evaluation details); finally, in Section 5, we set and schedule the tasks required to complete the project.
% To-do: rewrite the outline

@@

% Experiments
% - Document modelling: issue with pLSI and mixture unigrams~ 
% - Document classification: 
% -- Dimensionality reduction
% -- Comparison with SVM
\par The conducted experiments display an improvement in performance over the predecessor models and suggest the application domains of LDA-like models. The first experiment addresses document modelling. To be more specific, LDA has displayed balanced hidden topic proportions in the set of tested documents. That is, the tested documents have replicated the topic proportions of the training documents. The second experiment is directed towards the problem of document classification. The authors have addressed the \textit{dimensionality reduction} in order to surpass the performance of the support vector machine (SVM) models. To be more specific, the documents with reduced dimensionality would possess a lower amount of features. The results of the experiment indicate that the basic LDA model has successfully reduced the number of features. Also, it has displayed an improvement in accuracy compared to the SVM-like models.

@@

% Inference
% Variational methods are a deterministic alternavive for stochastic simulation
% Two variations: Kalman filtering and wavelet regression
\par The authors have used variational methods to approximate the inference of the posterior. It is claimed that stochastic simulation (sampling-based variance methods) would be unable to scale with larger data sets. Further, the authors discuss two methods used for the variational inference: Kalman Filtering and Wavelet Regression. The implementation details of these methods can be found in the `Dynamic Topic Modelling' article \cite{blei_2006}. Note that the implementation details are provided in a brief manner suggesting the need for supplementary literature.

@@

The initial dynamic topic modelling paper suggests inducing the time-series into the generative process of data sets. However, the way we capture the ions is done instantaneously for the whole corpus. Therefore, we are relaxing the assumption of time-series. Instead of that, we will use the dynamic topic modelling techniques to establish the relationship between the time-slices (even though we are not taking time into account, we will stick to the same terminology to preserve the consistency). For example, if we assume that the data is generated in a 1-dimensional manner, then time-slice $t$ would depend on its adjacent time-slices $t-1$ and $t+1$. However, since the previous illustration of the sample is 2-dimensional, there are more adjacent time-slices. This relationship is shown in Figure \ref{fig:2-d_generative_process} below.
\begin{figure}[H]
  \centering
  \includegraphics[width=0.5\textwidth]{2-d_generative_process}
  \caption{2-dimensional generative process.}
  \label{fig:2-d_generative_process}
\end{figure}
In the boundary case shown on the left, time-slice $t_{1,1}$ is related to three surrounding time-slices, whereas, in the general case indicated on the right side, time-slice $t_{2,2}$ relates to eight surrounding time-slices. Ultimately, the notion of dynamic topic modelling can be  preserved even upon relaxing the assumption of time-series.

@@

% Inference
% - Metropolis--Hastings
\par In Section 3, we have addressed a sampling-based inference method -- Gibbs sampling. However, for the dynamic topic models, the latter method has to be reconsidered due to the change from Dirichlet to Gaussian distributions. Therefore, during the implementation of the model, we will attempt to utilise another MCMC-based algorithm -- Metropolis--Hastings.
% - To-do: Analyse the details of the MH.

@@

% Intro
% - Spatial Smoothing
% - Two inference methods
% - Risk management
\par The developed technique for the spatial smoothing of the MSI data sets will be evaluated by measuring the improvement in performance. That is, the performance will be assessed by investigating the quality of the inferred patterns. Also, since we plan to develop two models with different inference methods, we will be able to assess the time performance with respect to the inference method used. Hence, we will be able to deduce which variation scales better with the size of the data sets. Furthermore, in this section, we will assess the possible risks; also, we will provide potential strategies for mitigating the risks.

% 
% - Generating synthetic data sets
% -- The improvement in the inference
\par The model will be evaluated using synthetic and real data sets. However, at the start, we will only use the synthetic data sets. Note that the performance will be measured using the following algorithm: (1) take two equivalent data sets; (2) apply the spatial smoothing on one of the data sets; (3) infer the underlying structure of the data; (4) check whether the applied smoothing has inferred the underlying structure better. By working with synthetic data sets, we have the control over inducing the underlying structure of the data. Hence, we will able to give a basis for measuring the performance of the applied spatial smoothing.

% Risks
% - Algorithm complexity (implementation difficulty)
% - Size of the data sets -> slow computation 
% - Obtaining real data
% - Uncertainty whether the model produces an improvement
\par Even though we have induced a strategy for the project's evaluation, the described process assumes that we can mitigate the following risks:
\begin{itemize}
  \item The implementation complexity of the inference algorithms;
  \item The access to the real data.
  \item The model's scalability in large data sets;
  \item The inability to measure the performance of the applied spatial smoothing.
\end{itemize}
The first three risks would slow down the project execution thus raise a concern of finishing the project on time, whereas the fourth one would prevent deriving the confirmation or the denial of the initial project's hypothesis -- `the degeneracies of MSI data sets can be detected using LDA-like models'.

% Risk mitigation
% - Algorithms
% -- Familiarising with them before the research
% -- Starting with the simpler one (Gibbs sampling)
\par The risk of complex (in terms of development) algorithms has been already mitigated by familiarising with basic sampling-based and optimisation-based inference models. That is, we have implemented the models utilising Gibbs sampling and variational inference. Also, note that the main model will firstly be developed using Gibbs sampling, as this method is easier to implement compared to the variational one. However, as previously discussed, we intend to exchange the use of the Gibbs sampling method by utilising the Metropolis--Hastings method.

@@

% - Real data can be obtained from the Web
\par The risk of the real data access might be partially resolved, as the University has a laboratory for the research in metabolomics. However, in case this options is not possible, we can obtain the data sets from the Web.

% - Uncertainty in the ground truth
% -- Generate data sets with underlying structure
% -- Assess the inference by using spatial and non-spatial models
\par Finally, the risk of the inability to measure the performance of the applied smoothing can impact the outcome of the project the most. That is, it would be difficult to reason about the improvement in the inference of the real data sets because we do not possess the full knowledge of the data sets' underlying structure. We will mitigate this issue by establishing a performance metric from the synthetic data sets.

% Evaluation/Experiments
% - Real data
\par By having the defined boundaries of the expected improvement in performance, we can attempt to tackle the real data sets. That is, the expected performance improvement would serve as an indicator whether the inferred real data deviates in the expected limit of improvement. To be more specific, the data sets with applied spatial smoothing and the data sets without applied spatial smoothing will display some differences in the inferred topics. Hence, we will check whether those differences meet the performance boundaries suggested by the inference of the synthetic data sets.

@@

% Intro
% - Tasks.
% - Schedule
% - Deliverables
\par In this section, we will outline the project milestones and deliverables. That is, we will outline the most important steps and provide the estimations in the form of the Gantt chart. Further, the deliverables will be induced from the project's objectives and hypothesis (these are addressed in Section 2). The deliverables will contain both successful and successful outcomes of the set hypothesis.

@@

% Tasks
% - Design
% -- Investigate the literature on the spatial smoothing application in MSI data sets
% -- Design the dynamic topic model
% - Implementation
% -- Implement the generative process of the dynamic topic modelling with applied spatial smoothing
% -- Implement both sampling-based and variational inference to the dynamic topic model
% - Evaluation
% -- Test the performance of the applied spatial smoothing
% - Documentation/Presentation
% -- Dissertation
% -- Reports
% -- Presentation
\par The high-level plan of the project's execution include the following milestones: design, implementation, evaluation, and documentation/presentation. To establish a basis for the design, we will continuously review the literature on spatial smoothing, MSI, and dynamic topic models. Then, the implementation will require the utilisation of spatial smoothing in the generative process of the dynamic topic model; also, for the inference, we will implement both: sampling-based and variational methods. Further, the evaluation will be conducted by assessing the performance of the developed models. Finally, the results will be documented in the forms of weekly reports, the dissertation, and the presentation.

% Dates
% - Non-continuous
% - Estimations
\par The outlined tasks will not be necessarily executed in a sequential manner. For some of the tasks, we will use the parallel approach as given in Figure \ref{fig:gantt} below.

@@

% - Successful/Unsuccessful
% - Model in Python 
% -- 2/3 objectives
% - Dissertation
% - Presentation
\par The deliverables of the project will depend on the outcome of the initial hypothesis -- `the degeneracies of MSI data sets can be detected using LDA-like models'. Nevertheless, in any of the project's outcomes, we will provide the following deliverables: the presentation; the dissertation; and the spatial smoothing model, which will be written in Python. However, the main project's contribution to the research in MSI will depend on the results: if the model displays an improvement in performance, then we could utilise the model in the pre-processing of MSI data sets; in the case of unsuccessful outcome, we would assess the techniques used and provide details on why the model does not work. Both outcomes would bring the motivation for further research in improving MSI data sets: the successful outcome would bring attention to the applicability of LDA-like models; the unsuccessful outcome would suggest the need of a different perspective required to tackle the issue of noise.
